Title

### 2020-11-25 documents/DraftReport.Rmd
Added intro outline, did some methods clean-up.

1. Briefly describe the issue, conclude with teaser sentence about what this project does
IPMs as an increasingly popular tool that can provide tangible benefits
Non-ID parameters, improved precision
Expanded spatio-temporal inference
These benefits have motivated interesting expansions and applications of this technique
Examples
First (?) joint analysis of mark-recovery and census data (Besbeas et al. 2002)
Application to species (bats) w/ demographic data deficiencies (Schaub et al. 2007)
Estimating hidden demographic processes (e.g., proportion of female lambs on the island and temporal variation) (Tavecchia et al. 2009)
“We experimentally evaluated the flexibility and limitations of the integrated approach by omitting portions of the data and comparing the results from these “incomplete” data sets with those obtained from the entire data set and from the analysis of the mark-resighting-recovery data only”
Inclusion of multivariate observations w/ estimation of transition probabilities within multi-state systems (McCrea et al. 2012)
Inclusion of environmental stochasticity (Oppel et al. 2014)
However, this growth in usage has occurred with relatively poor understanding of how to assess goodness of fit, and therefore less is also known about the bounds/limits of the IPM framework to estimate demographic parameters in situations where available data might be sparse. 
Study summary: In this study, we use simulated data to examine IPM performance across a range of scenarios where data informing key demographic parameters may be either sparse or missing.

2. With the expansion of the use of IPMs, few studies have robustly examined IPM performance through data simulation. 
In those that have used data simulation, the focus has been on:
Proof of concept (improved precision and reduced bias)
McCrea et al 2010, Chandler & Clark 2014, Schaub & Fletcher 2015, Bled et al. 2017, Ahrestani et al. 2017, Bowler et al. 2019
Model assumptions
Independence assumption (Abadi et al. 2010, Schaub & Fletcher 2015, Weegman et al. 2020)
MR assumptions (survival heterogeneity, marker-induced bias; Riecke et al. 2019)
None/few to date have looked at ability to return parameter values or power to detect trends when survival not informed by data or when a given data source is sparse, biased, or highly uncertain, but this is an interesting, complicated, and important topic
IPM performance is often context-specific: depends on life history, sample size, effect sizes, likelihood structures, study/inference goals, so what can we learn
Contributions of different datasets to the likelihood - helps us understand the underlying connections/mechanisms of the IPM, and to use this knowledge to think about IPM performance across life histories, survey designs, etc.

3. The “typical” IPM has gotten a lot of mileage, but can we push the limits?
The typical (survival and counts) has taught a lot, and been extended
Examples of a few cool extensions: spatially explicit IPMs, source-sink dynamics, environmental stochasticity
But wouldn’t it also be cool and open doors if we could apply this approach in situations where this “typical” data wasn’t available or was sparse? 
Situations where only know fecundity and counts
Situations where some of our data streams are biased and/or uncertain?
Best to worst: unbiased/low variance, unbiased/high variance, biased/low variance, biased/high variance
Sun et al. 2019 (adding sites vs. changing detection)
Simmonds et al. 2020; Farr et al. 2020 (similar to our goals but for SDMs)
We can’t confidently go in these directions without refining our understanding of the mechanisms underlying IPMs in these situations and where and when IPMs may not lead to improved precision or where they actually can’t estimate that unidentifiable parameter
This study fills that knowledge gap by X (restate purpose/approach) with the aim of improving our understanding of this important statistical approach, particularly for passerine bird life history

4. Why this study is useful for practitioners
Study design: minimizing cost/effort and maximizing usefulness
Surveys per week, hours per survey, number of sites, etc
Intentional survey design vs opportunistic/community science program data  
Sanderlin et al. 2014
Helping ecologists understand what is possible with available data, particularly given the tendency to maximize what you have 
Importance of unbiased estimates for conservation and management

@@

*Data simulation: life history*  
We simulated data for an avian species assuming life history stages and traits of a typical passerine. Our simulated hypothetical species has two age-classes: young-of-year (YOY) and adults. Each age class has a unique survival rate ($S_1$, $S_{ad}$) and adults produce new chicks at rate *f*, and those chicks survive to be YOY at rate $S_1$. Current YOY will also breed the following year to produce surviving chicks at rate $f \times S_1$. (Fig 1). We assume an annual breeding cycle, with females nesting once per year. The period from nest initiation to fledging is 30 days regardless of clutch size.

@@

*Count model*  
We simulated data such that the population is surveyed annually in a 15-year pre-breeding census. Count surveys are conducted three times per year in quick succession such that we assume the population is closed between secondary sampling periods. We use an N-mixture model to estimate annual abundance from these repeated counts, where we assume no individual is double counted and that individual detections are independent and occur with equal probability. The underlying state process represents the latent variables of stage-specific abundance at each occasion:

$$N_{1, t+1} \sim Poisson(N_{1,t}, S_{juv,t}, \frac{f_t}{2} + N_{ad,t}S_{ad,t},\frac{f_t}{2})$$  
$$N_{ad,t+1} \sim Binomial(N_{1,t} + N_{ad,t}, S_{ad,t})$$

where the number of juveniles ($N_1$) is estimated using a Poisson distribution with probability based on the sum of stage-specific individuals that survive and reproduce. A binomial distribution is used to estimate the number of adults based on the stage-specific survival and abundance in the previous occasion.

@@

$$y_t = (N_{tot, t}) + \epsilon$$
$$\epsilon \sim Normal(0, \sigma^2_y)$$

describes counts *y* at time *t* given given stage-specific abundance as binomially distributed with sample size $N_t$ and detection probability $p$, with normally distributed observation error $\epsilon$ estimated with mean = 0 with the prior for $sigma_y$ being U(0,10). 
<!-- from amanda: help - prior for observation error?  -->

The overall likelihood for this hierarchical state-space model can be described as the product of the state and observation processe models:
$$L_{SS}(\text{y}|\text{N}, \text{S_juv}, \text{S_ad}, \text{f}, \sigma^2_y) = L_O(\text{y}|\text{N},\sigma^2_y) \times L_S(\text{N}|\text{S_juv}, \text{S_ad}, \text{f})$$

@@

$$J_t \sim Poisson(\rho_t) $$
$$\rho_t = R_tf_t $$

where the number of nestlings counted $J$ in year *t* is modeled using a Poisson distribution with the product of the number of surveyed broods $R_t$ and fecundity $f_t$ at each time *t*. Fecundity is estimated using a uniform prior (0,20) and the likelihood of this daily nest survival model can be described as $L_P(\text{J}, \text{R}|\text{f})$.

@@

$$ z_{i,t}|z_{i,t-1} \sim bernoulli(z_{i,t-1} S_{1,ad}) $$

@@

$$ y_{i,t}|z_{i,t} \sim  bernoulli(z_{i,t} p)$$

describes observations *y* using a Bernouli distribution based on the state of the individual in the current occasion and detection probability *p*. Mean survival and detection probabilities are modeled on the probability scale using uniform priors, $ S_{1,ad} \sim uniform(0,1)$, $ p \sim uniform(0,1)$, respectively. The overall likelihood of the mark-resight model estimating survival probabilities is the product of the state and observation process models: $L_{SS}(\text{y}|\text{z}, \text{S_juv}, \text{S_ad}, \text{p}) = L_O(\text{y}|\text{z},\text{p})     \times L_S(\text{z}|\text{S_juv}, \text{S_ad})$.

@@

$$ \left[\begin{array}
{l}
N_{1,t+1} \sim Binomial(N_{1,t}S_{1,t}) \\
N_{ad,t+1} \sim Binomial(N_{ad,t}S_{ad,t}) \\
N_{Tot, t+1} = N_{1,t} + N_{ad,t}
\end{array}\right] 
$$

The joint likelihood for the integrated population model is the product of the likelihoods for the three subcomponent models. In instances when one dataset is removed, informative priors are used in place of the data to inform each of the relevant parameters (Table 1). 
Integration: leslie matrix/pop model; joint likelihood equations

@@

@@

@@

`r figs("plots", caption = "Scatter and boxplots of rain per site.")`

@@

`r tbls("rain_table", caption = "Table of rain by site.")`
```{r}

# kable(rain_table, align = 'c', format.args = list(big.mark = ","))

>>>>

1. Briefly describe the issue, conclude with teaser sentence about what this project does

- IPMs as an increasingly popular tool that can provide tangible benefits  
-- Non-ID parameters, improved precision  
-- Expanded spatio-temporal inference  
-- These benefits have motivated interesting expansions and applications of this technique  
--- Examples   
- Brief history of IPM development  
-- First (?) joint analysis of mark-recovery and census data (Besbeas et al. 2002)  
-- Application to species (bats) w/ demographic data deficiencies (Schaub et al. 2007)  
-- Estimating hidden demographic processes (e.g., proportion of female lambs on the island and temporal variation) (Tavecchia et al. 2009)  
*“We experimentally evaluated the flexibility and limitations of the integrated approach by omitting portions of the data and comparing the results from these “incomplete” data sets with those obtained from the entire data set and from the analysis of the mark-resighting-recovery data only”*  
-- Inclusion of multivariate observations w/ estimation of transition probabilities within multi-state systems (McCrea et al. 2012)  
-- Inclusion of environmental stochasticity (Oppel et al. 2014)  
- However, this growth in usage has occurred with relatively poor understanding of how to assess goodness of fit, and therefore less is also known about the bounds/limits of the IPM framework to estimate demographic parameters in situations where available data might be sparse. 
- Study summary: In this study, we use simulated data to examine IPM performance across a range of scenarios where data informing key demographic parameters may be either sparse or missing.

2. With the expansion of the use of IPMs, few studies have robustly examined IPM performance through data simulation. 
- In those that have used data simulation, the focus has been on:
-- Proof of concept (improved precision and reduced bias): McCrea et al 2010, Chandler & Clark 2014, Schaub & Fletcher 2015, Bled et al. 2017, Ahrestani et al. 2017, Bowler et al. 2019  
- Model assumptions  
-- Independence assumption (Abadi et al. 2010, Schaub & Fletcher 2015, Weegman et al. 2020)  
-- MR assumptions (survival heterogeneity, marker-induced bias; Riecke et al. 2019)  
- None/few to date have looked at ability to return parameter values or power to detect trends when survival not informed by data or when a given data source is sparse, biased, or highly uncertain, but this is an interesting, complicated, and important topic  
- IPM performance is often context-specific: depends on life history, sample size, effect sizes, likelihood structures, study/inference goals, so what can we learn  
- Contributions of different datasets to the likelihood - helps us understand the underlying connections/mechanisms of the IPM, and to use this knowledge to think about IPM performance across life histories, survey designs, etc.

3. The “typical” IPM has gotten a lot of mileage, but can we push the limits?  
- The typical (survival and counts) has taught a lot, and been extended  
- Examples of a few cool extensions: spatially explicit IPMs, source-sink dynamics, environmental stochasticity
- But wouldn’t it also be cool and open doors if we could apply this approach in situations where this “typical” data wasn’t available or was sparse?  
-- Situations where only know fecundity and counts  
-- Situations where some of our data streams are biased and/or uncertain?  
-- Best to worst: unbiased/low variance, unbiased/high variance, biased/low variance, biased/high variance
--- Sun et al. 2019 (adding sites vs. changing detection)  
--- Simmonds et al. 2020; Farr et al. 2020 (similar to our goals but for SDMs)  
- We can’t confidently go in these directions without refining our understanding of the mechanisms underlying IPMs in these situations and where and when IPMs may not lead to improved precision or where they actually can’t estimate that unidentifiable parameter. This study fills that knowledge gap by X (restate purpose/approach) with the aim of improving our understanding of this important statistical approach, particularly for passerine bird life history.

4. Why this study is useful for practitioners    
- Study design: minimizing cost/effort and maximizing usefulness
- Surveys per week, hours per survey, number of sites, etc  
- Intentional survey design vs opportunistic/community science program data: Sanderlin et al. 2014
- Helping ecologists understand what is possible with available data, particularly given the tendency to maximize what you have 
- Importance of unbiased estimates for conservation and management

@@

*Data simulation: life history*  
We simulated data for an avian species assuming life history stages and traits of a typical passerine. Our simulated hypothetical species has two age-classes: young-of-year (YOY) and adults. Each age class has a unique survival rate ($S_1$, $S_{ad}$) and adults produce new chicks at rate *f*, and those chicks survive to be YOY at rate $S_1$. Current YOY will also breed the following year to produce surviving chicks at rate $f \times S_1$ (Fig 1). We assume an annual breeding cycle, with females nesting once per year. The period from nest initiation to fledging is 30 days regardless of clutch size.

@@

*Count model*  
We simulated data such that the population is surveyed annually in a 15-year pre-breeding census. Count surveys are conducted three times per year in quick succession such that we assume the population is closed between secondary sampling periods. We use an N-mixture model (citations) to estimate annual abundance from these repeated counts, where we assume no individual is double counted and that individual detections are independent and occur with equal probability. The underlying state process represents the latent variables of stage-specific abundance at each occasion:

$$N_{1, t+1} \sim \text{Poisson}(N_{1,t}, S_{1,t}, \frac{f_t}{2} + N_{ad,t}S_{ad,t},\frac{f_t}{2})$$

$$N_{ad,t+1} \sim \text{Binomial}(N_{1,t} + N_{ad,t}, S_{ad,t})$$

where the number of juveniles ($N_1$) is estimated using a Poisson distribution with probability based on the sum of stage-specific individuals that survive ($S_{1,ad}$) and reproduce ($f_t$). A binomial distribution is used to estimate the number of adults based on the stage-specific survival and abundance in the previous occasion.

@@

$$y_t = N_{tot, t} + \epsilon$$
$$\epsilon \sim \text{Normal}(0, \sigma^2_y)$$

describes counts *y* at time *t* given given stage-specific abundance as binomially distributed with sample size $N_t$ and detection probability $p$, with normally distributed observation error $\epsilon$ estimated with mean = 0 with the prior for $\sigma_y$ being U(0,10). 
<!-- from amanda: help - prior for observation error?  -->

The overall likelihood for this hierarchical state-space model can be described as the product of the state and observation process models:
$$L_{SS}(\text{y}|\text{N}, S_1, S_{ad}, f, \sigma^2_y) = L_O(\text{y}|\text{N},\sigma^2_y) \times L_S(\text{N}|S_1, S_{ad}, f)$$

@@

$$J_t \sim \text{Poisson}(\rho_t) $$
$$\rho_t = R_tf_t $$

where the number of nestlings counted $J$ in year *t* is modeled using a Poisson distribution with the product of the number of surveyed broods $R_t$ and fecundity $f_t$ at each occasion. Fecundity is estimated using a uniform prior (0, 20) and the likelihood of this daily nest survival model can be described as $L_P(\text{J}, \text{R}|\text{f})$.

@@

$$ z_{i,t}|z_{i,t-1} \sim \text{Bernoulli}(z_{i,t-1} S_{1,ad}) $$

@@

$$ y_{i,t}|z_{i,t} \sim  \text{Bernoulli}(z_{i,t} p)$$

describes observations *y* using a Bernouli distribution based on the state of the individual in the current occasion and detection probability *p*. Mean survival and detection probabilities are modeled on the probability scale using uniform priors, $S_{1,ad} \sim \text{Uniform}(0,1)$ and $p \sim \text{Uniform}(0,1)$, respectively. The overall likelihood of the mark-resight model estimating survival probabilities is the product of the state and observation process models: $L_{SS}(\text{y}|\text{z}, ~S_1, ~S_{ad}, ~\text{p}) = L_O(\text{y}|\text{z},~\text{p})     \times L_S(\text{z}|S_1,~S_{ad})$.

@@

<!-- From amanda: help: not sure if we need something like this since we have the equations above? -->
<!-- $$ \left[\begin{array} -->
<!-- {l} -->
<!-- N_{1,t+1} \sim \text{Binomial}(N_{1,t},~S_{1,t}) \\ -->
<!-- N_{ad,t+1} \sim \text{Binomial}(N_{ad,t},~S_{ad,t}) \\ -->
<!-- N_{Tot, t+1} = N_{1,t} + N_{ad,t} -->
<!-- \end{array}\right]  -->
<!-- $$  -->

The joint likelihood for the integrated population model is the product of the likelihoods for the three subcomponent models. In instances when one dataset is removed, informative priors are used in place of the data to inform each of the relevant parameters (Table 1).

[IPM joint likelihood equation]

@@

Review project aim and main findings

Elaborate on interesting finding 1  
- Implications of not exploring this in an application?  
- Future areas of work?  
- Extension to more complex life history? Data scenarios?

Elaborate on interesting finding 2
- Implications of not exploring this in an application?  
- Future areas of work?  
- Extension to more complex life history? Data scenarios?

Broaden to implications of work, study design, existing IPMs

@@

Abadi, F., Gimenez, O., Arlettaz, R., Schaub, M., 2010. An assessment of integrated population models: bias, accuracy, and violation of the assumption of independence. Ecology 91, 7–14. https://doi.org/10.1890/08-2235.1

Ahrestani, F.S., Saracco, J.F., Sauer, J.R., Pardieck, K.L., Royle, J.A., 2017. An integrated population model for bird monitoring in North America. Ecol Appl 27, 916–924. https://doi.org/10.1002/eap.1493

Bled, F., Belant, J.L., Daele, L.J.V., Svoboda, N., Gustine, D., Hilderbrand, G., Barnes, V.G., n.d. Using multiple data types and integrated population models to improve our knowledge of apex predator population dynamics. Ecology and Evolution 7, 9531–9543. https://doi.org/10.1002/ece3.3469

Bowler, D.E., Nilsen, E.B., Bischof, R., O’Hara, R.B., Yu, T.T., Oo, T., Aung, M., Linnell, J.D.C., 2019. Integrating data from different survey types for population monitoring of an endangered species: the case of the Eld’s deer. Scientific Reports 9, 7766. https://doi.org/10.1038/s41598-019-44075-9

Chandler, R.B., Clark, J.D., 2014. Spatially explicit integrated population models. Methods in Ecology and Evolution 5, 1351–1360. https://doi.org/10.1111/2041-210X.12153

Farr, M.T., Green, D.S., Holekamp, K.E., Zipkin, E.F., n.d. Integrating distance sampling and presence-only data to estimate species abundance. Ecology n/a, e03204. https://doi.org/10.1002/ecy.3204

McCrea, R.S., Morgan, B.J.T., Gimenez, O., Besbeas, P., Lebreton, J.-D., Bregnballe, T., 2010. Multi-Site Integrated Population Modelling. Journal of Agricultural, Biological and Environmental Statistics 15, 539–561. https://doi.org/10.1007/s13253-010-0027-5

Riecke, T.V., Williams, P.J., Behnke, T.L., Gibson, D., Leach, A.G., Sedinger, B.S., Street, P.A., Sedinger, J.S., 2019. Integrated population models: model assumptions and inference. Methods Ecol Evol 2041–210X.13195. https://doi.org/10.1111/2041-210X.13195

Schaub, M., Fletcher, D., 2015. Estimating immigration using a Bayesian integrated population model: choice of 
parametrization and priors. Environ Ecol Stat 22, 535–549. https://doi.org/10.1007/s10651-015-0309-8

Simmonds, E.G., Jarvis, S.G., Henrys, P.A., Isaac, N.J.B., O’Hara, R.B., 2020. Is more data always better? A simulation study of benefits and limitations of integrated distribution models. Ecography 43, 1413–1422. https://doi.org/10.1111/ecog.05146

Sun, C.C., Royle, J.A., Fuller, A.K., 2019. Incorporating citizen science data in spatially explicit integrated population models. Ecology 100, e02777. https://doi.org/10.1002/ecy.2777

Tavecchia, G., Besbeas, P., Coulson, T., Morgan, B.J.T., Clutton‐Brock, T.H., 2009. Estimating Population Size and Hidden Demographic Parameters with State‐Space Modeling. The American Naturalist 173, 722–733. https://doi.org/10.1086/598499

Weegman, M.D., Arnold, T.W., Clark, R.G., Schaub, M., 2020. Partial and complete dependency among data sets has minimal consequence on estimates from integrated population models. Ecological Applications. https://doi.org/10.1002/eap.2258

@@

`r figs("plots", caption = "Plots.")`

@@

`r tbls("rain_table", caption = "Table of X, Y, Z.")`
```{r}

# kable(table, align = 'c', format.args = list(big.mark = ","))
